On the computational complexity of detecting possibilistic locality

### 2017-06-19 paper.tex
Changed Mansfield citation to article where complexity is considered
Boring commit: unifying Mansfield Refs- no need to summarise

In 1991, Avis \emph{et al} \cite{Avis2004} demonstrated that determining whether or not a given table of these operational predictions for a Bell experiment had a local hidden variable explanation was \textbf{NP}-hard, building on work by Pitowsky \cite{Pito1991}. In this paper, we consider a more extreme, possibilistic, manifestation of nonlocality; one in which it is not only true that we cannot account for the specific experimental probabilities with a local hidden variable model, but that we cannot even account for the possibilities. In doing so, we will identify the minimal operational requirements for this problem to remain \textbf{NP}-hard, filling in the final gap in a categorisation started by Mansfield and Fritz \cite{Mans2011,Mans2016} and continued by Abramsky, Gottlob and Kolaitis \cite{Abra2013}. This nonlocality is a feature only of the table of possibilities at hand, with no dependence on the source of the correlations. Accordingly, the first sections of this paper will focus only on the decision problem at hand, and no knowledge about how such a table might be calculated within the formalism of quantum mechanics is required. In the final section, we turn our focus to those data tables accessible within quantum theory, and demonstrate how we can embed a large set of \textsc{3-SAT} problems into a quantum formalism.

@@

%Need to add:  What does it meant to be quantumly accessible? What's the complexity of finding out if one is? %Define measurement subrow and subcolumn, define "context"
%use consistent notation for the measurement setting selection
\section{The Locality Decision Problem}

@@

Question: do there exist a set of functions $T_i$, such that the $T_i$ are deterministic, and $T=\sum_i T_i$?
\end{defn}
Tihs represents a two-party nonlocality scenario in which Alice has access to measurements with $m$ outcomes, and Bob has access to measurements with $l$ outcomes. We are given a possibility table and asked whether or not it displays nonlocality. In general, we will take $|A|=|B|=n$ to maintain a single scaling factor; in fact this change is without loss of generality. In the case where $l=m$, we will denote this problem merely $l$-\textsc{PossLoc}. While this is not essential, we will also require that the data tables obey the possibilistic no-signalling principle; this merely streamlines the wording of the theorems. 
\begin{thm} \cite{Mans2011}
2-\textsc{PossLoc} is in \textbf{P}.
\end{thm}
\begin{proof}
This proof proceeds by showing that the only kind of possibilistic nonlocality that can occur in the two-outcome case is the one originally identified by Hardy \cite{Hard1993}; that is to say, a feature of the following kind, in which a blank space can be substituted either for a 0 or for a 1:
\begin{equation*}
\begin{array}{c| c c | c c |}
&&&&\\ \hline
&1 & & 0 & \\
&& & & \\ \hline
&0& & &  \\
&& & & 0 \\ \hline
\end{array}
\end{equation*}
We can quickly see that the highlighted one is impossible to extend to a ``deterministic grid'', \emph{i.e.} no single deterministic hidden variable can account for that possibility without also introducing an event that is impossible in the diagram. Since the appearance of this structure is equivalent to possibilistic nonlocality in this case, an algorithm to determine whether or not a given data table is possibilistically local or nonlocal reduces to checking each possible set of four contexts of this type, of which there are $n^2(n-1)^2/4$. Since checking for the appearance of such structures is possible in constant time, our algorithm runs in $O(n^4)$.
\end{proof}
However, this argument is specific to the case in which we have measurements with at most two outcomes on each side. As an illustration that there are quantumly-accessible nonlocal data tables that are not reducible to a fine-graining of a Hardy paradox, we will now demonstrate a novel possibilistic nonlocality scenario in the (2,3) case. This will be a generalisation of Hardy's proof of Bell's theorem.

\begin{figure}
\begin{center}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=0.8cm,y=0.8cm]
\clip(-5.,-4.) rectangle (5.,5.);
\draw(0.,0.) circle (3.2cm);
\draw (-2.5074154767005594,3.116547388890871)-- (2.5262219891238353,3.101322695507066);
\draw (-0.6234859067055393,3.9511093789136704)-- (3.70768134486122,-1.501032659527465);
\draw (0.6202539684631545,3.951618024886225)-- (-3.6765801200722317,-1.575677194316671);
\draw [dash pattern=on 2pt off 2pt] (-2.5074154767005594,3.116547388890871)-- (-3.6765801200722317,-1.575677194316671);
\draw [dash pattern=on 2pt off 2pt] (-3.6765801200722317,-1.575677194316671)-- (3.70768134486122,-1.501032659527465);
\draw [dash pattern=on 2pt off 2pt] (-3.6765801200722317,-1.575677194316671)-- (2.5262219891238353,3.101322695507066);
\draw [dash pattern=on 2pt off 2pt] (-2.5074154767005594,3.116547388890871)-- (3.70768134486122,-1.501032659527465);
\draw [dash pattern=on 2pt off 2pt] (3.70768134486122,-1.501032659527465)-- (2.5262219891238353,3.101322695507066);
\begin{scriptsize}
\draw [fill=uuuuuu] (0.,0.) circle (1.5pt);
\draw [fill=xdxdff] (-2.5074154767005594,3.116547388890871) circle (1.5pt);
\draw[color=xdxdff] (-2.8168692003229885,3.7112493895076946) node {$x$};
\draw [fill=xdxdff] (2.5262219891238353,3.101322695507066) circle (1.5pt);
\draw[color=xdxdff] (2.7461074833053605,3.5513937376792932) node {$y$};
\draw [fill=xdxdff] (-0.6234859067055393,3.9511093789136704) circle (1.5pt);
\draw[color=xdxdff] (-0.802687987285138,4.47855651828402) node {$a$};
\draw [fill=xdxdff] (3.70768134486122,-1.501032659527465) circle (1.5pt);
\draw[color=xdxdff] (4.024952697932568,-1.2762469475384213) node {$A$};
\draw [fill=xdxdff] (0.6202539684631545,3.951618024886225) circle (1.5pt);
\draw[color=xdxdff] (0.8598107917302307,4.4146142575526595) node {$b$};
\draw [fill=xdxdff] (-3.6765801200722317,-1.575677194316671) circle (1.5pt);
\draw[color=xdxdff] (-4.127685545315876,-1.244275817172741) node {$B$};
\end{scriptsize}
\end{tikzpicture}
\end{center}
\caption{Points on the Bloch sphere to which Alice's ensemble can be steered, revealing the generalised Hardy paradox.}
\label{steer}
\end{figure}

Constructing the possibility table for this steering scenario reveals the following structure in a subset of the table:
\begin{equation*}
\begin{array}{ c  | c c | c c c| c c c|}\
  & b&b^\perp& A^\perp &B^\perp&x^\perp& A^\perp &B^\perp&y^\perp \\ \hline
  a &1&1&1&1&1&1&1&1\\ 
  A &1&1&0&1&1&0&1&1\\ \hline
  b &1&0&1&1&1&1&1&1\\ 
 B  &1&1&1&0&1&1&0&1\\ \hline
 x &1&1&1&1&0&1&1&1\\ 
 y &1&1&1&1&1&1&1&0\\ \hline
\end{array}
\end{equation*}
We now present this paradox minimally, demonstrating that it is neither a Hardy paradox, nor a fine-graining of a Hardy paradox. Therefore, the Hardy paradox is not universal even for quantumly-accessible possibilistic nonlocality as long as one party has access to three-outcome measurements.

\begin{equation*}
\begin{array}{ c | c c | c c c| c c c|}
&&&&&&&& \\ \hline
&1&&0&&&0&&\\
   &&&&&&&&\\ \hline
   &0&&&&&&&\\ 
  &&&&0&&&0&\\ \hline
  &&&&&0&&&\\ 
  &&&&&&&&0\\ \hline
\end{array}
\end{equation*}
We see that in some sense this can be thought of as a generalisation of Hardy's paradox; the third measurement row acts to convert the two right-hand measurement columns into a single effective column identical to that of the standard Hardy paradox; we see that at least one of the two right-hand columns must have its completion of the highlighted 1 in a subcolumn other than the rightmost subcolumn, causing a Hardy paradox. We note also that this family of no-signalling distributions has a supremum value of the \emph{paradoxical probability} (the maximum probability of a measurement witnessing a Hardy paradox) of $\nicefrac{1}{2}$. This is a large violation compared to other known situations with a fixed number of outcomes, a fixed quantum dimension, and a fixed number of measurements: the paradoxical probability for Hardy's own construction is roughly 0.09. This probability can be met by scenarios with two parties with access to $k$ different 2-outcome measurements, or by a generalised no-signalling theory \cite{Mans2016}. This is because the overlap between $\ket{A}$ and $\ket{b^\perp}$ can be made arbitrarily large, and the reduced state of the system can be made arbitrarily close to the completely mixed state.

>>>>

In 1991, Avis \emph{et al} \cite{Avis2004} demonstrated that determining whether or not a given table of these operational predictions for a Bell experiment had a local hidden variable explanation was \textbf{NP}-hard, building on work by Pitowsky \cite{Pito1991}. In this paper, we consider a more extreme, possibilistic, manifestation of nonlocality; one in which it is not only true that we cannot account for the specific experimental probabilities with a local hidden variable model, but that we cannot even account for the possibilities. In doing so, we will identify the minimal operational requirements for this problem to remain \textbf{NP}-hard, filling in the final gap in a categorisation started by Mansfield and Fritz \cite{Mans2011,Mans2017} and continued by Abramsky, Gottlob and Kolaitis \cite{Abra2013}. This nonlocality is a feature only of the table of possibilities at hand, with no dependence on the source of the correlations. Accordingly, the first sections of this paper will focus only on the decision problem at hand, and no knowledge about how such a table might be calculated within the formalism of quantum mechanics is required. In the final section, we turn our focus to those data tables accessible within quantum theory, and demonstrate how we can embed a large set of \textsc{3-SAT} problems into a quantum formalism.

@@

%Need to add:  What does it meant to be quantumly accessible? What's the complexity of finding out if one is? %Define measurement subrow and subcolumn, define "context"
%use consistent notation for the measurement setting selection

%Mention how such tables can otherwise appear eg Abramsky- otherwise explain why in letter to referee.

\section{The Locality Decision Problem}

@@

Question: do there exist a set of functions $T_i$, such that the $T_i$ are deterministic, and $T=\sum_i T_i$?
\end{defn}
Tihs represents a two-party nonlocality scenario in which Alice has access to measurements with $m$ outcomes, and Bob has access to measurements with $l$ outcomes. We are given a possibility table and asked whether or not it displays nonlocality. In general, we will take $|A|=|B|=n$ to maintain a single scaling factor; in fact this change is without loss of generality. In the case where $l=m$, we will denote this problem merely $l$-\textsc{PossLoc}. While this is not essential, we will also require that the data tables obey the possibilistic no-signalling principle; this merely streamlines the wording of the theorems. 
\begin{thm} \cite{Mans2017}
2-\textsc{PossLoc} is in \textbf{P}.
\end{thm}
\begin{proof}
This proof proceeds by showing that the only kind of possibilistic nonlocality that can occur in the two-outcome case is the one originally identified by Hardy \cite{Hard1993}; that is to say, a feature of the following kind, in which a blank space can be substituted either for a 0 or for a 1:
\begin{equation*}
\begin{array}{c| c c | c c |}
&&&&\\ \hline
&1 & & 0 & \\
&& & & \\ \hline
&0& & &  \\
&& & & 0 \\ \hline
\end{array}
\end{equation*}
We can quickly see that the highlighted one is impossible to extend to a ``deterministic grid'', \emph{i.e.} no single deterministic hidden variable can account for that possibility without also introducing an event that is impossible in the diagram. Since the appearance of this structure is equivalent to possibilistic nonlocality in this case, an algorithm to determine whether or not a given data table is possibilistically local or nonlocal reduces to checking each possible set of four contexts of this type, of which there are $n^2(n-1)^2/4$. Since checking for the appearance of such structures is possible in constant time, our algorithm runs in $O(n^4)$.
\end{proof}
However, this argument is specific to the case in which we have measurements with at most two outcomes on each side. As an illustration that there are quantumly-accessible nonlocal data tables that are not reducible to a fine-graining of a Hardy paradox, we will now demonstrate a novel possibilistic nonlocality scenario in the (2,3) case. This will be a generalisation of Hardy's proof of Bell's theorem.

\begin{figure}
\begin{center}
\begin{tikzpicture}[line cap=round,line join=round,>=triangle 45,x=0.8cm,y=0.8cm]
\clip(-5.,-4.) rectangle (5.,5.);
\draw(0.,0.) circle (3.2cm);
\draw (-2.5074154767005594,3.116547388890871)-- (2.5262219891238353,3.101322695507066);
\draw (-0.6234859067055393,3.9511093789136704)-- (3.70768134486122,-1.501032659527465);
\draw (0.6202539684631545,3.951618024886225)-- (-3.6765801200722317,-1.575677194316671);
\draw [dash pattern=on 2pt off 2pt] (-2.5074154767005594,3.116547388890871)-- (-3.6765801200722317,-1.575677194316671);
\draw [dash pattern=on 2pt off 2pt] (-3.6765801200722317,-1.575677194316671)-- (3.70768134486122,-1.501032659527465);
\draw [dash pattern=on 2pt off 2pt] (-3.6765801200722317,-1.575677194316671)-- (2.5262219891238353,3.101322695507066);
\draw [dash pattern=on 2pt off 2pt] (-2.5074154767005594,3.116547388890871)-- (3.70768134486122,-1.501032659527465);
\draw [dash pattern=on 2pt off 2pt] (3.70768134486122,-1.501032659527465)-- (2.5262219891238353,3.101322695507066);
\begin{scriptsize}
\draw [fill=uuuuuu] (0.,0.) circle (1.5pt);
\draw [fill=xdxdff] (-2.5074154767005594,3.116547388890871) circle (1.5pt);
\draw[color=xdxdff] (-2.8168692003229885,3.7112493895076946) node {$x$};
\draw [fill=xdxdff] (2.5262219891238353,3.101322695507066) circle (1.5pt);
\draw[color=xdxdff] (2.7461074833053605,3.5513937376792932) node {$y$};
\draw [fill=xdxdff] (-0.6234859067055393,3.9511093789136704) circle (1.5pt);
\draw[color=xdxdff] (-0.802687987285138,4.47855651828402) node {$a$};
\draw [fill=xdxdff] (3.70768134486122,-1.501032659527465) circle (1.5pt);
\draw[color=xdxdff] (4.024952697932568,-1.2762469475384213) node {$A$};
\draw [fill=xdxdff] (0.6202539684631545,3.951618024886225) circle (1.5pt);
\draw[color=xdxdff] (0.8598107917302307,4.4146142575526595) node {$b$};
\draw [fill=xdxdff] (-3.6765801200722317,-1.575677194316671) circle (1.5pt);
\draw[color=xdxdff] (-4.127685545315876,-1.244275817172741) node {$B$};
\end{scriptsize}
\end{tikzpicture}
\end{center}
\caption{Points on the Bloch sphere to which Alice's ensemble can be steered, revealing the generalised Hardy paradox.}
\label{steer}
\end{figure}
%Need to define what's going on in this figure! Then show how it gives rise to these possibilities. Give combinatorial condition.
Constructing the possibility table for this steering scenario reveals the following structure in a subset of the table:
\begin{equation*}
\begin{array}{ c  | c c | c c c| c c c|}\
  & b&b^\perp& A^\perp &B^\perp&x^\perp& A^\perp &B^\perp&y^\perp \\ \hline
  a &1&1&1&1&1&1&1&1\\ 
  A &1&1&0&1&1&0&1&1\\ \hline
  b &1&0&1&1&1&1&1&1\\ 
 B  &1&1&1&0&1&1&0&1\\ \hline
 x &1&1&1&1&0&1&1&1\\ 
 y &1&1&1&1&1&1&1&0\\ \hline
\end{array}
\end{equation*}
We now present this paradox minimally, demonstrating that it is neither a Hardy paradox, nor a fine-graining of a Hardy paradox. Therefore, the Hardy paradox is not universal even for quantumly-accessible possibilistic nonlocality as long as one party has access to three-outcome measurements.

\begin{equation*}
\begin{array}{ c | c c | c c c| c c c|}
&&&&&&&& \\ \hline
&1&&0&&&0&&\\
   &&&&&&&&\\ \hline
   &0&&&&&&&\\ 
  &&&&0&&&0&\\ \hline
  &&&&&0&&&\\ 
  &&&&&&&&0\\ \hline
\end{array}
\end{equation*}
We see that in some sense this can be thought of as a generalisation of Hardy's paradox; the third measurement row acts to convert the two right-hand measurement columns into a single effective column identical to that of the standard Hardy paradox; we see that at least one of the two right-hand columns must have its completion of the highlighted 1 in a subcolumn other than the rightmost subcolumn, causing a Hardy paradox. We note also that this family of no-signalling distributions has a supremum value of the \emph{paradoxical probability} (the maximum probability of a measurement witnessing a Hardy paradox) of $\nicefrac{1}{2}$. This is a large violation compared to other known situations with a fixed number of outcomes, a fixed quantum dimension, and a fixed number of measurements: the paradoxical probability for Hardy's own construction is roughly 0.09. This probability can be met by scenarios with two parties with access to $k$ different 2-outcome measurements, or by a generalised no-signalling theory \cite{Mans2017}. This is because the overlap between $\ket{A}$ and $\ket{b^\perp}$ can be made arbitrarily large, and the reduced state of the system can be made arbitrarily close to the completely mixed state.
